Applying Multi-Core Model Checking and Maximum Satisfiability to Hardware-Software Partitioning

### 2015-09-09 Docs/Paper/LaTex/template.tex
refactor abstract, introduction, and background

\begin{document}
\title{Applying Multi-Core Model Checking to Hardware-Software Partitioning in Embedded Systems}
\author{Alessandro Trindade\and Hussama Ismail\and Renato Degelo\and Edilson Galv\~ao\and Helder Silva \and Lucas Cordeiro}

@@

\begin{abstract}
We present an alternative approach to solve the hardware (HW) and software (SW) partitioning problem, which uses Bounded Model Checking (BMC) based on Satisfiability Modulo Theories (SMT) in conjunction with a multi-core support using Open Multi-Processing. In a nutshell, the multi-core SMT-based BMC approach allows initializing many verification instances based on the number of available processing cores. Each instance checks for a different optimum value until the optimization problem is satisfied. We implement and evaluate our algorithms on top of the Efficient SMT-Based Context-Bounded Model Checker (ESBMC). We also integrate the $\nu$Z tool into ESBMC, which is a state-of-art optimization tool based on SMT. We compare all our approaches to another state-of-the-art optimization tool (Matlab). Experimental results show that no single optimization tool solves all HW-SW partitioning benchmarks; however, Matlab and $\nu$Z are the most efficient ones to solve HW-SW partitioning problems, although ESBMC had a significant performance improvement in particular cases.
\end{abstract}

@@

Nowadays, with the strong development of embedded systems, the design phase plays an important role. At early stages, the design is split into separated flows: hardware and software. The partitioning decision process, which deals with decisions upon which parts of the application have to be designed in hardware (HW) and which in software (SW), must be supported by any well-structured methodology. If not, this leads to a number of issues (design flow interruptions, redesigns, and undesired iterations) which affects the overall development process, the quality, and the lifecycle of the final system. Starting at the 1990s, intensive research was performed, and several approaches proposed, as shown in ~\cite{Arato2003} and ~\cite{Mann2007}.

In any HW and SW design of complex systems, more time is spent on verification than on construction ~\cite{Baier2008}. Formal methods based on model checking offer great potential to obtain a more effective and faster verification in the design process. Software (and systems) may be viewed as mathematical objects with behavior that is, in principle, well determined. This makes it possible to specify software using mathematical logic, which constitutes the intended (correct) behavior. Then, one can try to give a formal proof or otherwise establish that the software meets its specification ~\cite{Clarke2009}. Research in formal methods has led to the development of very promising verification techniques, which facilitate the early detection of errors. Model-based verification techniques use models that describe the possible system behavior in a mathematically precise and unambiguous manner. The system models are accompanied by algorithms that systematically explore all the states of the system model.

In~\cite{Trindade2015,Trindade2016,Trindade2014} was shown that it is possible to use Bounded Model Checking (BMC) based on Satisfiability Modulo Theories (SMT) to perform HW-SW partitioning in embedded systems. The present work extends those studies since there is a substantial improvement in terms of the SMT-based verification method, which has been extended with a multi-core parallel- and binary-search approaches, as well as, the integration of the MaxSMT solver $\nu$Z into ESBMC, which is a state-of-art optimization tool based on SMT~\cite{Bjorner2015}. Multi-core processors have been used in all segments of industry to implement high-performance computing~\cite{Wu2014}. In particular, hardware platforms, together with multi-processing platforms, have allowed verification algorithms to distribute tasks executions across multiple processors, which generate an increase in performance if compared to single-core solution~\cite{Holzmann11}. However, most algorithms for software verification still disregard the limitations of the CMOS technology and focus their solution on a single-core approach~\cite{Cordeiro2012,Clarke2004,Armando2009,Ganai2006}.

Here, we exploit the availability of multi-core processors; in particular, SMT-based verification methods are applied to the HW-SW partition problem in three different ways using a multi-core ESBMC approach with OpenMP: ESBMC-SS using a sequential-search (SS), ESBMC-PS using a parallel-search (PS), and ESBMC-PB using a binary-search (BS). Experimental results are compared to ILP (integral linear programming), GA (generic algorithms) in a multi-core version, and also to $\nu$Z, which supports only a single-core approach~\cite{Bjorner2015}. The ILP and GA algorithms are implemented with the optimization toolbox of Matlab~\cite{OpenMP1998}, while $\nu$Z is a built-in tool to the SMT solver Z3 in which its main function is to check the satisfiability of logical formulas based on objective functions. All multi-core ESBMC approaches with OpenMP, together with $\nu$Z, are implemented with the Efficient SMT-based Context-Bounded Model Checker (ESBMC) tool~\cite{Cordeiro2012}.

\textbf{Contributions.} The main contribution of the present study is to show that it is possible to take advantage of an SMT-based BMC tool in a multi-core architecture to solve optimization problems. Experimental results show that multi-core model-checking techniques can be effective, in particular cases, to find the optimal solution of the HW-SW partitioning problem using an SMT-based BMC approach. Additionally, we integrate the MaxSMT solver $\nu$Z into an off-the-shelf BMC tool that is typically used for software verification. To the best of our knowledge, this is the first work to use a multi-core SMT-based verification and a MaxSMT solver to check for HW-SW partitioning problems in embedded systems.

@@

The HW-SW partitioning problem is typically represented as a set of constraints and an objective function in linear programming. In this section, we describe the linear programming problem and present related tools that are used to model and solve the HW-SW partitioning problem.

@@

Optimization is the act of obtaining the best result (i.e., the optimal solution) under given circumstances~\cite{Rao2009}. In the design, construction, and maintenance of any engineering system, engineers have to make many technological and managerial decisions at several stages. The ultimate goal of all such decisions is either to minimize the effort required or to maximize the desired benefit. Because the effort required or the benefit desired in any practical situation can be expressed as a function of certain decision variables, optimization can be defined as the process of finding the conditions that give the maximum or minimum value of a function~\cite{Rao2009}.

There is no single method available for solving all optimization problems efficiently~\cite{Rao2009}. The most well-known technique is linear programming, which is an method applicable for the solution of problems in which the objective function and the constraints appear as linear functions of the decision variables. A particular case of linear programming is ILP, in which the variables can assume just integer values. Eq.~\ref{linear-programming-problem} shows a typical linear programming problem, where $A$ and $b$ are vectors or matrixes that describe the constraints
\begin{equation}
\label{linear-programming-problem}
  minf^t x \: such \; that  = 
  \begin{cases}
    A.x \leq b, \\ 
    Aeq.x = beq, \\ 
    x \geq 0.
  \end{cases}
\end{equation}

@@

In this study, ESBMC has been used as a BMC tool to solve HW-SW partitioning problems~\cite{Cordeiro2012}. According to Cordeiro {\it et al.}~\cite{Cordeiro2012},\cite{Cordeiro2011}, SMT-based model checking can be used to verify single- and multi-threaded software in embedded systems. In Ramalho {\it et al.}~\cite{Ramalho2013}, ESBMC can also be used to model check C$++$ software based on SMT solvers. In Trindade {\it et al.}~\cite{Trindade2015},\cite{Trindade2014}, it was shown that it is possible to use ESBMC also as an optimization tool.

In particular, there are two directives in C/C++ that can be used to guide a model checker to solve an optimization problem: ASSUME and ASSERT. The directive ASSUME is responsible for ensuring the compliance of constraints (software costs), and the directive ASSERT controls the halt condition or code violation (minimum hardware cost). Then, with some C/C++ code, it is possible to guide ESBMC to solve optimization problems.

@@

Fig.~\ref{ESBMC-Architecture} shows the current ESBMC architecture, which consists of the C/C$++$ parser, GOTO Program, GOTO Symex, and SMT solver ~\cite{Ramalho2013}. In particular, ESBMC compiles C/C$++$ code into equivalent GOTO\hyp{}programs ({\it i.e.}, control-flow graphs) using a gcc-compliant style. GOTO-programs can then be processed by the symbolic execution engine, called GOTO Symex, where two recursive functions compute the constraints ($C$) and properties ($P$); finally it generates two sets of equations (i.e.,\:$C \land \neg P$ ), which are checked for satisfiability by an SMT solver. The main factor for ESBMC to use only a single-core relies on its back-end (i.e., SMT Solver). Currently, the SMT solvers supported by ESBMC are: Z3~\cite{DeMoura2008}, Boolector~\cite{Brummayer2009}, MathSAT~\cite{Barrett2011}, CVC4~\cite{Bozzano2005}, and Yices~\cite{Dutertre2014}. Most of them do provide neither multi-threaded support nor a parallel version to solve the generated SMT equations.
%
\begin{figure}[ht]
	\centering
  \includegraphics[scale=0.9]{Image/esbmc-arch-new.pdf} 
	\caption{ESBMC Architecture}
	\label{ESBMC-Architecture}
\end{figure}

@@

In the above example, the for loop is executed in parallel. Each iteration of the loop is executed in a separated thread; and each thread may use an idle processor. There is also a way to specify critical regions, which is a code block that is guaranteed to be executed by a single thread at a time. To create a critical region, the ``\#pragma omp critical'' directive should be used.
%----------------------------------------------
\subsection{Solving Optimization Problems with $\nu$Z}
\label{Optimization-with-Vz}
%----------------------------------------------

SMT decides the satisfiability of first-order formulas using a combination of different background theories and thus generalizes propositional satisfiability by supporting uninterpreted functions, linear and nonlinear arithmetic, bitvectors, tuples, arrays, and other decidable first-order theories. In this study, the SMT solver Z3 is used to check for the satisfiability of formulas generated from the HW-SW partitioning problem~\cite{Bjorner2014}. In particular, we exploit the use of $\nu$Z, which is implemented on top of the SMT solver Z3, in order to solve optimization problems; $\nu$Z base function is to optimize objective functions, which formulate optimized criteria, within the logical context of constraints.~$\nu$Z also includes an incremental version of the Maximum Resiliency (MaxRes) \cite{Federica2008} in order to achieve Maximum Satisfiability (MaxSAT) \cite{NarodytskaN} and a Simplex to solve numbers without defined patterns. In $\nu$Z, MaxSAT is responsible for the restrictions, while OptSMT optimizes linear arithmetic objectives~\cite{Bjorner2015}. In summary, $\nu$Z provides three main functions that extend Z3 for solving optimization problems, which are: \textit{maximize}, \textit{minimize}, and \textit{assert-soft}.

@@

The $\nu$Z architecture is shown in Fig.~\ref{vZ-Architecture}. Initially, the SMT formula with objectives is converted to $0-1$ constraints, which leads to a Pseudo-Boolean Optimization (PBO)~\cite{Barth1995,Vasco2005}. If there are many objective functions, $\nu$Z invokes OptSAT for arithmetic or MaxSAT for soft constraints. For constraints using real values, $\nu$Z combines linear arithmetic objectives and uses only one instance of OptSMT. When ``soft constrains'' is used in the mode `` lexicographic'', $\nu$Z invokes MaxSAT using multiple calls for its engine.
%
\begin{figure}[ht]
	\centering
  \includegraphics[width=0.49\textwidth, height=95px]{Image/vzArch.png} 
	\caption{$\nu$Z Architecture~\cite{Bjorner2015}}
	\label{vZ-Architecture}
\end{figure}

>>>>

\begin{document}
\title{Applying Multi-Core Model Checking and Maximum Satisfiability to Hardware-Software Partitioning}
\author{Alessandro Trindade\and Hussama Ismail\and Renato Degelo\and Edilson Galv\~ao\and Helder Silva \and Lucas Cordeiro}

@@

\begin{abstract}
We present an alternative approach to solve the hardware (HW) and software (SW) partitioning problem, which uses Bounded Model Checking (BMC) based on Satisfiability Modulo Theories (SMT) in conjunction with a multi-core support using Open Multi-Processing. In a nutshell, the multi-core SMT-based BMC approach allows initializing many verification instances based on the number of available processing cores. Each instance checks for a different optimum value until the optimization problem is satisfied. We implement our algorithms on top of the Efficient SMT-Based Context-Bounded Model Checker (ESBMC) and also integrate the maximum satisfiability solver $\nu$Z tool into ESBMC. We compare all our approaches to another state-of-the-art optimization tool (Matlab). Experimental results show that there is no single optimization tool to solve all HW-SW partitioning benchmarks; however, Matlab and ESBMC-$\nu$Z are the most efficient ones to solve HW-SW partitioning problems, although multi-core ESBMC had a significant performance improvement in particular cases.
\end{abstract}

@@

Nowadays, with the strong development of embedded systems, the design phase plays an important role. At early stages, the design is split into separated flows: hardware and software. The partitioning decision process, which deals with decisions upon which parts of the application have to be designed in hardware (HW) and which in software (SW), must be supported by any well-structured methodology. If not, this leads to a number of issues (design flow interruptions, redesigns, and undesired iterations) which affect the overall development process, the quality, and the lifecycle of the final system.

\comment{In any HW and SW design of complex systems, more time is spent on verification than on construction ~\cite{Baier2008}. Formal methods based on model checking offer great potential to obtain a more effective and faster verification in the design process. Software (and systems) may be viewed as mathematical objects with behavior that is, in principle, well determined. This makes it possible to specify software using mathematical logic, which constitutes the intended (correct) behavior. Then, one can try to give a formal proof or otherwise establish that the software meets its specification ~\cite{Clarke2009}. Research in formal methods has led to the development of very promising verification techniques, which facilitate the early detection of errors. Model-based verification techniques use models that describe the possible system behavior in a mathematically precise and unambiguous manner. The system models are accompanied by algorithms that systematically explore all the states of the system model.}

Starting at the 1990s, intensive research was performed in HW-SW partitioning, and several approaches proposed, as shown in ~\cite{Arato2003} and~\cite{Mann2007}. In~\cite{Trindade2015,Trindade2016,Trindade2014} was shown that it is possible to use Bounded Model Checking (BMC) based on Satisfiability Modulo Theories (SMT), implemented in a tool called Efficient SMT-Based Context-Bounded Model Checker (ESBMC), in order to perform HW-SW partitioning in embedded systems. The present work extends those studies since there is a substantial improvement in terms of the SMT-based verification method, which has been extended with a multi-core parallel- and binary-search approaches, as well as, the integration of the Maximum SMT (MaxSMT) solver $\nu$Z into ESBMC, which is a state-of-art optimization tool based on SMT~\cite{Bjorner2015}.

\comment{Multi-core processors have been used in all segments of industry to implement high-performance computing~\cite{Wu2014}. In particular, hardware platforms, together with multi-processing platforms, have allowed verification algorithms to distribute tasks executions across multiple processors, which generate an increase in performance if compared to single-core solution~\cite{Holzmann11}. However, most algorithms for software verification still disregard the limitations of the CMOS technology and focus their solution on a single-core approach~\cite{Cordeiro2012,Clarke2004,Armando2009,Ganai2006}.}

Here, we exploit the availability of multi-core processors; in particular, SMT-based verification methods are applied to the HW-SW partition problem in three different ways using a multi-core ESBMC approach with OpenMP: ESBMC-SS using a sequential-search (SS), ESBMC-PS using a parallel-search (PS), and ESBMC-PB using a binary-search (BS). Experimental results are compared to ILP (integral linear programming), GA (generic algorithms) in a multi-core version, and also to $\nu$Z, which supports only a single-core approach~\cite{Bjorner2015}. The ILP and GA algorithms are implemented with the optimization toolbox of Matlab~\cite{OpenMP1998}, while $\nu$Z is a built-in tool to the SMT solver Z3. All multi-core ESBMC approaches, together with $\nu$Z, are implemented with the Efficient SMT-based Context-Bounded Model Checker (ESBMC) tool~\cite{Cordeiro2012}.

% in which its main function is to check the satisfiability of logical formulas based on objective functions

\textbf{Contributions.} The main contribution of the present study is to describe and evaluate a comprehensive SMT-based BMC approach in a multi-core architecture applied to solve HW-SW optimization problems. Additionally, we integrate the MaxSMT solver $\nu$Z into an off-the-shelf BMC tool, which is typically used for software verification, in order to formulate and solve optimization problems within the logical context of constraints. Experimental results show that multi-core model-checking techniques can be effective, in particular cases, to find the optimal solution of the HW-SW partitioning problem using an SMT-based BMC approach. Although there is no single tool for efficiently solving all HW-SW partitioning problems, we show that the MaxSMT solver $\nu$Z is faster than other state-of-the-art optimization tools for small- and medium-size optimization problems.  To the best of our knowledge, this is the first work to use a multi-core SMT-based verification and a MaxSMT solver to check for HW-SW partitioning problems in embedded systems.

@@

The HW-SW partitioning problem is typically represented as a set of constraints and an objective function in linear programming. We describe the linear programming problem and present related tools that are used to model and solve the HW-SW partitioning problem.

@@

Optimization is the act of obtaining the best result ({\it i.e.}, the optimal solution) under given circumstances~\cite{Rao2009}. 
\comment{In the design, construction, and maintenance of any engineering system, engineers have to make many technological and managerial decisions at several stages. The ultimate goal of all such decisions is either to minimize the effort required or to maximize the desired benefit. Because the effort required or the benefit desired in any practical situation can be expressed as a function of certain decision variables, optimization can be defined as the process of finding the conditions that give the maximum or minimum value of a function~\cite{Rao2009}.}
There is no single method available for efficiently solving all optimization problems~\cite{Rao2009}. The most well-known technique is linear programming, which is an method applicable for the solution of problems in which the objective function and the constraints appear as linear functions of the decision variables. A particular case of linear programming is ILP, in which the variables can assume just integer values. Eq.~\ref{linear-programming-problem} shows a typical linear programming problem, where $A$ and $b$ are vectors or matrixes that describe the constraints
\begin{equation}
\label{linear-programming-problem}
  minf^t x \: such \; that  = 
  \begin{cases}
    A.x \leq b, \\ 
    Aeq.x = beq, \\ 
    x \geq 0.
  \end{cases}
\end{equation}

@@

In this study, ESBMC has been used as a BMC tool to solve HW-SW partitioning problems~\cite{Cordeiro2012}. 
\comment{According to Cordeiro {\it et al.}~\cite{Cordeiro2012},\cite{Cordeiro2011}, SMT-based model checking can be used to verify single- and multi-threaded software in embedded systems. In Ramalho {\it et al.}~\cite{Ramalho2013}, ESBMC can also be used to model check C$++$ software based on SMT solvers. In Trindade {\it et al.}~\cite{Trindade2015},\cite{Trindade2014}, it was shown that it is possible to use ESBMC also as an optimization tool.}
In particular, there are two directives in ESBMC that can be used to guide a model checker to solve an optimization problem: ASSUME and ASSERT. The directive ASSUME is responsible for ensuring the compliance of constraints (software costs), and the directive ASSERT controls the halt condition (minimum hardware cost). Then, with some C/C++ code, it is possible to guide ESBMC to solve optimization problems.

@@

Fig.~\ref{ESBMC-Architecture} shows the current ESBMC architecture, which consists of the C/C$++$ parser, GOTO Program, GOTO Symex, and SMT solver ~\cite{Ramalho2013}. In particular, ESBMC compiles C/C$++$ code into equivalent GOTO\hyp{}programs ({\it i.e.}, control-flow graphs) using a gcc-compliant style. GOTO-programs can then be processed by the symbolic execution engine, called GOTO Symex, where two recursive functions compute the constraints ($C$) and properties ($P$); finally, it generates two sets of equations (i.e.,\:$C \land \neg P$ ), which are checked for satisfiability by an SMT solver. The main factor for ESBMC to use only a single-core relies on its back-end (i.e., SMT Solver). Currently, the SMT solvers supported by ESBMC are: Z3~\cite{DeMoura2008}, Boolector~\cite{Brummayer2009}, MathSAT~\cite{Barrett2011}, CVC4~\cite{Bozzano2005}, and Yices~\cite{Dutertre2014}. Most of them do provide neither multi-threaded support nor a parallel version to solve the generated SMT equations.
%
\begin{figure}[ht]
	\centering
  \includegraphics[scale=0.9]{Image/esbmc-arch-new.pdf} 
	\caption{ESBMC Architecture}
	\label{ESBMC-Architecture}
\end{figure}

@@

In the above example, the for loop is executed in parallel. Each iteration of the loop is executed in a separated thread; and each thread may use an idle processor. There is also a way to specify critical regions, which is a code block that is guaranteed to be executed by a single thread at a time. To create a critical region, the ``\#pragma omp critical'' directive is used.

%----------------------------------------------
\subsection{Solving Optimization Problems with $\nu$Z}
\label{Optimization-with-Vz}
%----------------------------------------------

\comment{SMT decides the satisfiability of first-order formulas using a combination of different background theories and thus generalizes propositional satisfiability by supporting uninterpreted functions, linear and nonlinear arithmetic, bitvectors, tuples, arrays, and other decidable first-order theories.}

In this study, the SMT solver Z3 is used to check for the satisfiability of formulas generated from the HW-SW partitioning problem~\cite{Bjorner2014}. In particular, we exploit the use of MaxSMT solver $\nu$Z, which is implemented on top of the SMT solver Z3, in order to solve optimization problems; $\nu$Z base function is to optimize objective functions, which formulate optimized criteria, within the logical context of constraints.~$\nu$Z also includes an incremental version of the Maximum Resiliency (MaxRes)~\cite{Federica2008}, in order to achieve Maximum Satisfiability (MaxSAT)~\cite{NarodytskaN} and a Simplex to solve numbers without defined patterns.

In $\nu$Z, MaxSAT is responsible for the restrictions, while OptSMT optimizes linear arithmetic objectives~\cite{Bjorner2015}. In summary, $\nu$Z provides three main functions that extend Z3 for solving optimization problems, which are: \textit{maximize}, \textit{minimize}, and \textit{assert-soft}.

@@

Fig.~\ref{vZ-Architecture} shows the $\nu$Z architecture. Initially, the SMT formula with objectives is converted to $0-1$ constraints, which leads to a Pseudo-Boolean Optimization (PBO)~\cite{Barth1995,Vasco2005}. If there are many objective functions, $\nu$Z invokes OptSAT for arithmetic or MaxSAT for soft constraints. For constraints using real values, $\nu$Z combines linear arithmetic objectives and uses only one instance of OptSMT. When ``soft constrains'' is used in the mode `` lexicographic'', $\nu$Z invokes MaxSAT using multiple calls for its engine.
%
\begin{figure}[ht]
	\centering
  \includegraphics[width=0.49\textwidth, height=95px]{Image/vzArch.png} 
	\caption{$\nu$Z Architecture extracted from~\cite{Bjorner2015}}
	\label{vZ-Architecture}
\end{figure}
